---
title: "STAT 412: Case Study 1"
author: "Your name(s)"
date: "Due: October 15, 2025"
output: pdf_document
---

DELETE ALL THE TEXT PROVIDED BY BECKY BEFORE SUBMITTING!

```{r setup, message = F, eval = FALSE}
# if you need packages, load them here

# the following line of code sets the size of the figures when knitting for all code chunks. 
# You can also change the figure size in each specific chunk's header
knitr::opts_chunk$set(fig.height = 3, fig.width = 5)

# load your data here
# Once you've set your file path, change the R chunk header to eval = TRUE
marriage <- read_csv()

```


## Model description

Clearly state your model (Address questions 1 and 2). I encourage you to type up your model in math notation using Latex. For example:

$Y_{1}, \ldots, Y_{n} |\theta \overset{iid}{\sim} f(y | \theta)$

If you do not know how, that's okay! Just let me know and I will be happy to help or point you towards some resources. 




## Implementation

Clearly state how you will implement your model. (Address question 3.) You
should provide enough detail such that someone else could implement your analysis.

## Gibbs sampler


```{r gibbs_sampler}
# code for your gibbs sampler should go here
# remember to only keep necessary code for submission!
```

## Results

### Diagnostics

First do some diagnostics to see if your chain has converged

```{r}

```

### Findings

Provide results with interpretation (Address question 4). You should alternate between R code and text!

```{r}

```

## Conclusion

Give a conclusion that answers the research question (Address question 5).
