---
title: "STAT 412: Case Study 2"
author: "Your name(s)"
date: "Due: November 12, 2025"
output: pdf_document
---

DELETE ALL THE TEXT PROVIDED BY BECKY BEFORE SUBMITTING!

```{r setup, message = F, eval = FALSE}
# if you need packages, load them here

# the following line of code sets the size of the figures when knitting for all code chunks. 
# You can also change the figure size in each specific chunk's header
knitr::opts_chunk$set(fig.height = 3, fig.width = 5)

# load your data here
# Once you've set your file path, change the R chunk header to eval = TRUE
marriage <- read_csv()
```

## Introduction

Briefly recap the research task at hand.

## Model description

Clearly state your model (address questions 1 and 2). I encourage you to type up your model in math notation using Latex. For example:

$$Y_{1}, \ldots, Y_{n} |\theta \overset{iid}{\sim} f(y | \theta)$$
### Methods for answering research question

Clearly describe how you will use the model to address the research question. This includes providing some derivations or explanation for the reader (address question 3).

## Implementation

Clearly state how you will implement your model (address question 4). You should provide enough detail such that someone else could implement your analysis.

## MCMC code

```{r mcmc-function, eval = F}
# DELETE BECKY's COMMENTS BEFORE SUBMITTING!


# if you'd like to make a function, it looks like this:
my_function <- function(x, y, n, ...){
  # code you'd like to run in your function
  a <- sum(y)/n
  b <- sum(x)/n
  
  # this is a way to return multiple values from a function. The following code will return the a and b created above as objects called v1 and v2.
  ret <- list(v1 = a, v2 = b,...)
  return(ret)
}

# then, if you'd like to call your function and access the returned value, you could do the following (delete before running your own code):
mcmc1 <- my_function(x = x1, y = y1, n = n1,...)
v1_1 <- mcmc1$v1
```

```{r mcmc}
# code for your MCMC should go here
# remember to only keep display code necessary for submission!
# I highly recommend making a your sampler into a function and putting it in the chunk above when you're ready
```

### Diagnostics

```{r}

```


## Results


### Findings

Provide results with interpretation (address question 5). You should alternate between R code and text!

```{r}

```

## Conclusion

Give a conclusion that answers the research question and expands on the findings (address question 6).
